---
title: "Getting started with syncR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with syncR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(syncR)
```

## Overview

syncR is the integrator and coordinator of the Circadia Lab R ecosystem.

> *syncR::sync() — pulling everything into alignment, just like the SCN does.*

The package pulls three streams of participant data into a single unified,
participant-indexed database:

| Source   | Package     | Data type                                     |
|----------|-------------|-----------------------------------------------|
| `tallie` | **tallieR** | Sociodemographics, one-time questionnaires    |
| `slumb`  | **slumbR**  | Sleep diary entries, subjective sleep quality |
| `zeit`   | **zeitR**   | Actigraphy metrics, circadian parameters      |

## Basic usage

```{r, eval = FALSE}
library(syncR)

db <- sync(
  tallie = tallieR::export(),
  slumb  = slumbR::export(),
  zeit   = zeitR::export()
)
```

The result is a tidy, participant-indexed tibble with one row per participant
per night (where applicable), with columns from all supplied sources.

## Joining strategies

By default `sync()` performs a left join anchored on the first supplied source.
Use the `join` argument to change this:

```{r, eval = FALSE}
# Only participants present in all three sources
db_complete <- sync(
  tallie = tallieR::export(),
  slumb  = slumbR::export(),
  zeit   = zeitR::export(),
  join   = "inner"
)

# All participants from any source, NAs where data is absent
db_all <- sync(
  tallie = tallieR::export(),
  slumb  = slumbR::export(),
  zeit   = zeitR::export(),
  join   = "full"
)
```

## Custom participant ID column

If your data uses a different participant identifier column name, pass it via
`id_col`:

```{r, eval = FALSE}
db <- sync(
  tallie = my_tallie_data,
  zeit   = my_zeit_data,
  id_col = "pid"
)
```
